---
title: "Untitled"
---

```{r}
# Phylogenetics
library(ape)
library(treeio)

# tidyverse
library(tibble)
library(dplyr)
library(readr)
library(purrr)

# Utils
library(igraph)
library(geneplast)
```

# Baixando espécies do string

```{r}
download.file(
  url = "https://stringdb-downloads.org/download/species.v12.0.txt",
  destfile = here::here("data-raw/download/species.v12.0.txt")
)
```

# Carregando espécies do string

```{r}
string_species <- readr::read_tsv(
  here::here("data-raw/download/species.v12.0.txt"),
  skip = 1,
  col_names = c(
    "taxid",
    "string_type",
    "string_name",
    "ncbi_official_name",
    "domain"
  ),
  col_types = cols_only(
    taxid = "c",
    string_name = "c",
    domain = "c"
  )
)

taxa_of_interest <- string_species |> filter(domain == "Eukaryotes")
```

# Downloading NCBI Taxonomy

```{r}
options(timeout=300)

download.file(
  url = "http://ftp.ncbi.nlm.nih.gov/pub/taxonomy/taxdump.tar.gz",
  destfile = here::here("data-raw/download/taxdump.tar.gz")
)

untar("data-raw/download/taxdump.tar.gz", exdir = "data-raw/download/taxdump")
```

# Loading NCBI Taxonomy

```{r}
# These .dmp files are tricky to read, the following read_delims are very hacky
# ncbi_merged_ids represents a dictionary mapping old ids to new ones
ncbi_merged_ids <- readr::read_delim(
  here::here("data-raw/download/taxdump/merged.dmp"),
  delim = "|",
  trim_ws = TRUE,
  col_names = c("taxid","new_taxid"),
  col_types = cols_only(
    taxid = "c",
    new_taxid = "c"
  )
)

ncbi_updated_ids_lookup <- tibble::deframe(ncbi_merged_ids)

# ncbi_edgelist represents the NCBI Taxonomy tree
ncbi_edgelist <- readr::read_delim(
  here::here("data-raw/download/taxdump/nodes.dmp"),
  skip = 1,
  delim = "|",
  trim_ws = TRUE,
  col_names = c("n1","n2","rank"),
  col_types = "ccc"
)

ncbi_taxon_names <- readr::read_delim(
  here::here("data-raw/download/taxdump/names.dmp"),
  delim = "|",
  trim_ws = TRUE,
  col_names = c("name","ncbi_name","type"),
  col_types = "cc-c"
)
```

# Updating STRING taxon IDs

Some organisms taxon IDs are outdated in STRING. We must update them to work with the most recent NCBI Taxonomy data.

```{r}
taxa_of_interest_updated <-
  taxa_of_interest |>
  dplyr::left_join(ncbi_merged_ids) |>
  dplyr::mutate(new_taxid = dplyr::coalesce(new_taxid, taxid))
```

# Carrega, limpa e atualiza newick do TimeTree

```{r}
timetree_tree <- ape::read.tree(
  file = here::here("data-raw/download/TTOL v5.1 Species Tree - Internal NCBI ID Labelled.nwk")
)

# We cannot build a working tree with multiple empty labels
# Therefore we assign unlabelled nodes an arbitrary label. eg: "fix_123"
unlabelled_nodes_indexes <- timetree_tree$node.label == ""
number_of_unlabelled_nodes <- sum(unlabelled_nodes_indexes)
timetree_tree$node.label[unlabelled_nodes_indexes] <- paste0("fix_", 1:number_of_unlabelled_nodes)

# It is much easier to work with the tree as an edgelist dataframe
# One way to obtain the df is by converting it into an igraph obj, then df
timetree_graph <- ape::as.igraph.phylo(timetree_tree, directed = TRUE)
timetree_edgelist <- igraph::as_data_frame(timetree_graph, what = "edges")

# Removing single quotes
timetree_edgelist$from <- gsub("'", '', timetree_edgelist$from)
timetree_edgelist$to <- gsub("'", '', timetree_edgelist$to)

# Updating taxon ids
temp_timetree_edgelist_from <- ncbi_updated_ids_lookup[timetree_edgelist$from]
temp_timetree_edgelist_to <- ncbi_updated_ids_lookup[timetree_edgelist$to]

timetree_edgelist_updated <- timetree_edgelist

timetree_edgelist_updated$from <- dplyr::coalesce(temp_timetree_edgelist_from, timetree_edgelist$from)
timetree_edgelist_updated$to <- dplyr::coalesce(temp_timetree_edgelist_from, timetree_edgelist$to)
```

# Aqui descobrimos quais taxa estão faltando no TimeTree (ou seja, serão enxertados à força). Note que essa busca é feita tanto nas folhas quanto nos vértices internos da árvore.

```{r}
taxa_of_interest_not_in_timetree <- ! taxa_of_interest_updated$new_taxid %in% c(timetree_edgelist_updated$from, timetree_edgelist_updated$to)

taxa_of_interest_not_in_timetree <- taxa_of_interest_updated[taxa_of_interest_not_in_timetree,]
```

# Transformando edgelist NCBI Taxonomy em grafo e depois em árvore

Uma forma fácil de obter uma árvore a partir de uma edgelist é convertendo-a primeiro em objeto igraph, e depois em árvore.

```{r}
# Creates a dictionary that maps taxids to official names
ncbi_scientific_names <-
  ncbi_taxon_names |>
  filter(type == "scientific name") |>
  dplyr::select(name, ncbi_name)

ncbi_scientific_names_lookup <- tibble::deframe(ncbi_scientific_names)

eukaryota_taxon_id <- subset(ncbi_taxon_names, ncbi_name == "Eukaryota", "name", drop = TRUE)

ncbi_graph <- igraph::graph_from_data_frame(
  d = ncbi_edgelist[,2:1],
  directed = TRUE
)

# Releasing memory
rm(ncbi_edgelist, ncbi_merged_ids)
gc()

ncbi_tree <- treeio::as.phylo(ncbi_graph)
```

# Quais taxa de interesse **não** estão presentes no TimeTree? Esses são os vértices que precisam ser enxertados.

```{r}
taxids_not_in_timetree <- taxa_of_interest_not_in_timetree$new_taxid
```

# Para cada taxon faltante, percorrer cada vértice ancestral perguntando se ele existe no TimeTree

```{r}
ncbi_graph_vids <- igraph::V(ncbi_graph) |> as.numeric() |> as.character()

ncbi_graph_vnames <- igraph::V(ncbi_graph)$name

ncbi_graph_vid_to_vname <- setNames(ncbi_graph_vnames, ncbi_graph_vids)

"%i missing taxa. Looking for the closest available taxa up in the hierarchy. This might take a while..." |> sprintf(length(taxids_not_in_timetree)) |> print()
  
timetree_edgelist_patch <- purrr::map_df(
  .x = taxids_not_in_timetree, 
  .f = function(current_taxid) {
    root_vertex_id <- as.numeric(igraph::V(ncbi_graph)[[current_taxid]])
    
    found_ancestor <- NULL
    
    igraph::dfs(
      graph = ncbi_graph,
      root = root_vertex_id,
      mode = "in",
      unreachable = FALSE,
      order = FALSE,
      in.callback = function(graph, current_vertex, extra) {
        
        vertex_id <- as.character(current_vertex[["vid"]])
        vertex_taxid <- ncbi_graph_vid_to_vname[vertex_id]
        
        ancestor_is_in_timetree <- vertex_taxid %in% timetree_edgelist_updated$from
        
        if(ancestor_is_in_timetree) {
          found_ancestor <<- vertex_taxid
          return(TRUE)
        }
        
        return(FALSE)
      }
    )
    
    data.frame(from = found_ancestor, to = current_taxid)
  },
  .progress = TRUE
)
```

# Enxertando ramos novos na árvore do TimeTree

```{r}
# "Grafting" missing vertices to the TimeTree edgelist
timetree_edgelist_patched <- rbind(timetree_edgelist_updated, timetree_edgelist_patch)

# It can happen that, after forceful grafting, some edges are duplicated or link to themselves
# This is not allowed, therefore we remove such occurrences
timetree_edgelist_patched %<>%
  unique |>
  dplyr::filter(from != to)

# Creating an igraph from the grafted tree
timetree_graph_patched <- igraph::graph_from_data_frame(timetree_edgelist_patched, directed = TRUE)
```

# Constraining everything to Eukaryotes

This whole process works much better for eukaryotes. The following block extracts the eukaryotic subtree from the complete tree for just the taxa of interest. Any non-eukaryotes are removed after it.

```{r}
eukaryote_vertex_in_patched_timetree <- igraph::V(timetree_graph_patched)[eukaryota_taxon_id]

leaves_of_interest_in_patched_timetree <- igraph::V(timetree_graph_patched)[taxa_of_interest_updated$new_taxid]

# Traversing the TimeTree graph from the Eukaryota vertex towards the taxa of interest
eukaryote_paths_in_patched_timetree <- igraph::shortest_paths(
  graph = timetree_graph_patched,
  from = eukaryote_vertex_in_patched_timetree,
  to = leaves_of_interest_in_patched_timetree,
  mode = "out"
)$vpath

eukaryote_vertices_in_patched_timetree <- eukaryote_paths_in_patched_timetree |> unlist() |> unique()

timetree_eukaryote_subgraph <- igraph::induced_subgraph(
  graph = timetree_graph_patched,
  vids = eukaryote_vertices_in_patched_timetree,
  impl = "create_from_scratch"
)

timetree_tree_patched <- treeio::as.phylo(timetree_eukaryote_subgraph)
```

```{r  fig.height=70, fig.width=8.5}
timetree_tree_patched <- geneplast:::rotatePhyloTree(timetree_tree_patched, "9606")

tip_came_from_patch <- timetree_tree_patched$tip.label %in% timetree_edgelist_patch$to

timetree_tree_patched$tip.color <- ifelse(tip_came_from_patch, yes = "#FF0000", no = "#000000")
timetree_tree_patched$tip.label <- ncbi_scientific_names_lookup[timetree_tree_patched$tip.label]
timetree_tree_patched$edge.length <- NULL

pdf(file = here::here("hybrid_tree_2023_2.pdf"), height = 70, width = 10)

plot(
  timetree_tree_patched,
  type       = "cladogram",
  tip.color  = timetree_tree_patched$tip.color,
  edge.color = "#bbbbbb",
  cex        = 0.3,
  edge.width = 0.1,
  no.margin  = T,
  direction  = "rightwards"
)

dev.off()
```
